dispRity demo

### 2015-10-01 doc/dispRity-tutorial.Rnw
Renamed Rnw file

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

This a quick demo to go through the beta version of the \dispRity package (v.0.1.0).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref} \hlstd{=} \hlstr{"release"}\hlstd{)}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}
\end{kframe}
\end{knitrout}

Note that we use the \texttt{release} branch here which is version 0.1.0.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
Future version will just require the specific function but this version (0.1.0) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
Additionally, I warmly recommend to install the packages \texttt{geomorph} or/and \texttt{Claddis} if you want to try this demo with your very own ordinated morphometric or cladistic data.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Loading demo and the package data}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
\hlcom{## An ordinated matrix with the tips only}
\hlkwd{data}\hlstd{(BeckLee_mat50)}
\hlkwd{dim}\hlstd{(BeckLee_mat50)}
\end{alltt}
\begin{verbatim}
## [1] 50 48
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
\end{alltt}
\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}
\begin{alltt}
\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
\hlkwd{data}\hlstd{(BeckLee_mat99)}
\hlkwd{dim}\hlstd{(BeckLee_mat99)}
\end{alltt}
\begin{verbatim}
## [1] 99 97
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##                  [,1]       [,2]       [,3]       [,4]        [,5]
## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
\end{verbatim}
\begin{alltt}
\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##            [,1]      [,2]       [,3]       [,4]       [,5]
## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
\end{verbatim}
\begin{alltt}
\hlcom{## A list of first and last occurrences data for some fossils}
\hlkwd{data}\hlstd{(BeckLee_ages)}
\hlkwd{head}\hlstd{(BeckLee_ages)}
\end{alltt}
\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}
\begin{alltt}
\hlcom{## And a phylogeny:}
\hlkwd{data}\hlstd{(BeckLee_tree)}
\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.8\linewidth]{figure/minimal-data-1}

}

\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## My ordinated matrix with tips only}
\hlstd{MyData_mat50} \hlkwb{<-} \hlstd{BeckLee_mat50}

\hlcom{## My ordinated matrix with tips and nodes}
\hlstd{MyData_mat99} \hlkwb{<-} \hlstd{BeckLee_mat99}

\hlcom{## My first and last occurence data}
\hlstd{MyData_ages} \hlkwb{<-} \hlstd{BeckLee_ages}

\hlcom{## My phylogenetic tree}
\hlstd{MyData_tree} \hlkwb{<-} \hlstd{BeckLee_tree}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Splitting the data}
\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(MyData_mat99, MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
\hlcom{## Bootstrapping the data}
\hlstd{bootstrapped_data} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(sliced_data,} \hlnum{100}\hlstd{)}

\hlcom{## Calculating disparity}
\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}

\hlcom{## Summarising the results}
\hlkwd{summary}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
##      series  n  mean  2.5%   25%   75% 97.5%
## 1 133.51104  3 16.67  0.00 15.51 18.67 24.35
## 2 100.13328 14 51.48 45.03 49.29 53.62 56.54
## 3  66.75552 23 64.77 60.09 63.26 66.34 69.58
## 4  33.37776 11 42.41 35.87 40.94 44.40 46.98
## 5         0 10 40.20 34.46 38.38 42.46 44.57
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.5\linewidth]{figure/minimal-Quick_go-1}

}

\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{str}\hlstd{(sum_of_ranges)}
\hlcom{## That's a 2512 lines output!}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Which class is the sum_of_ranges object?}
\hlkwd{class}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlcom{## We can summarise it using the S3 method print.dispRity}
\hlstd{sum_of_ranges}
\end{alltt}
\begin{verbatim}
## Disparity measurements across 5 series for 99 taxa. 
## Series:
## 133.51104, 100.13328, 66.75552, 33.37776, 0.
## Disparity calculated as: sum ranges for 97 dimensions.
## Data was split using continuous method.
## Data was bootstrapped 100 times, using the full bootstrap method.
\end{verbatim}
\begin{alltt}
\hlcom{## Some attentive ape package users will note the similarities with a phylo object...}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
\hlcom{## All the descendant of these node are crown and all the ancestors are stem}
\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(MyData_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(MyData_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}

\hlcom{## We then have to feed this information in a data frame with one column}
\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}

\hlcom{## We then can use the customised series function}
\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(MyData_mat50, factors)}

\hlcom{## This created a dispRity object containing two series: crown and stem}
\hlkwd{class}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlstd{crown_stem}
\end{alltt}
\begin{verbatim}
## 2 custom series for 50 taxa. 
## Series:
## crown, stem.
\end{verbatim}
\begin{alltt}
\hlcom{## This object contains three elements}
\hlkwd{names}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "data"   "taxa"   "series"
\end{verbatim}
\begin{alltt}
\hlcom{## With "data" being the list sub-matrices}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
\end{alltt}
\begin{verbatim}
## List of 2
##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
##   .. ..$ : NULL
##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
##   .. ..$ : NULL
\end{verbatim}
\begin{alltt}
\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
\end{alltt}
\begin{verbatim}
##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
\end{verbatim}
\begin{alltt}
\hlcom{## and finally, "series" containing information on the series type (custom) and names}
\hlcom{## (crown and stem)}
\hlstd{crown_stem}\hlopt{$}\hlstd{series}
\end{alltt}
\begin{verbatim}
## [1] "custom" "crown"  "stem"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Let's generate three time bins containing the data present every 40 Ma}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{alltt}
\hlcom{## Note that the function provides a warning saying that tips where single}
\hlcom{## points in time (no FAD/LAD information). We can fix that by adding the}
\hlcom{## age data for the taxa that have some longer occurrence spans.}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
\hlcom{## To entirely avoid the warning we could collect the occurrence span data}
\hlcom{## for all the taxa but that's not necessary. The function automatically}
\hlcom{## assumes no occurrence span time (i.e. single points in time) for all taxa}
\hlcom{## by default.}

\hlstd{time_bins}
\end{alltt}
\begin{verbatim}
## 3 discrete series for 50 taxa. 
## Series:
## 120-80, 80-40, 40-0.
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
## 3 discrete series for 50 taxa. 
## Series:
## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Let's generate four time slices every 40 Ma assuming a gradual evolution model}
\hlstd{time_slices} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),}
    \hlkwc{FADLAD} \hlstd{= MyData_ages)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
\hlstd{time_slices}
\end{alltt}
\begin{verbatim}
## 4 continuous series for 99 taxa. 
## Series:
## 120, 80, 40, 0.
\end{verbatim}
\begin{alltt}
\hlcom{## Note that in the same way as for the discrete method, we can also}
\hlcom{## automatically generate the slices}
\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
    \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{4}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
## 4 continuous series for 99 taxa. 
## Series:
## 133.51104, 89.00736, 44.50368, 0.
\end{verbatim}
\begin{alltt}
\hlcom{## But that gives the same silly results (slicing every 44.50368 Ma)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{boot.matrix}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50)}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 1000 times, using the full bootstrap method.
\end{verbatim}
\begin{alltt}
\hlcom{## As we can wee, the output is also a dispRity object that is summarised}
\hlcom{## automatically, and gives information on the data as well as the number of}
\hlcom{## bootstraps and the bootstraps methods.}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Bootstrapping with the single bootstrap method}
\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{boot.type} \hlstd{=} \hlstr{"single"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 1000 times, using the single bootstrap method.
\end{verbatim}
\begin{alltt}
\hlcom{## Bootstrapping with the full rarefaction}
\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 20 times, using the full bootstrap method.
## Data was fully rarefied (down to 3 taxa).
\end{verbatim}
\begin{alltt}
\hlcom{## Or with a set number of rarefaction levels}
\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{6}\hlopt{:}\hlnum{8}\hlstd{,}\hlnum{3}\hlstd{))}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 20 times, using the full bootstrap method.
## Data was rarefied with a maximum of 6, 7, 8 and 3 taxa.
\end{verbatim}
\begin{alltt}
\hlcom{## And removing the last axis (default)}
\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 1000 times, using the full bootstrap method. 
## The 6 last axis have been removed from the original data.
\end{verbatim}
\begin{alltt}
\hlcom{## Or with a fix value (50%)}
\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{0.5}\hlstd{)}
\end{alltt}
\begin{verbatim}
## Bootstrapped ordinated matrix with 50 taxa. 
## 1 unnamed series.
## Data was bootstrapped 1000 times, using the full bootstrap method. 
## The 35 last axis have been removed from the original data.
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Let's do a full rarefaction on the crown/stem data for later}
\hlstd{crown_stemBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(crown_stem,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{## And just bootstrapping for the time binning/slicing}
\hlstd{time_binsBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_bins)}
\hlstd{time_slicesBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_slices)}

>>>>

\documentclass{article}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }

\begin{document}
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
options(formatR.arrow=TRUE,width=90)
@

@@

This a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

@@

<<installation, eval=FALSE>>=
install.packages("devtools")
library(devtools)
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
@

Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
Future version will just require the specific function but this version (\Sexpr{packageVersion("dispRity")}) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
Additionally, I warmly recommend to install the packages \texttt{geomorph} or/and \texttt{Claddis} if you want to try this demo with your very own ordinated morphometric or cladistic data.

@@

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading demo and the package data
library(dispRity)

## An ordinated matrix with the tips only
data(BeckLee_mat50)
dim(BeckLee_mat50)
head(BeckLee_mat50[,1:5])

## An ordinated matrix with the tips and the nodes (named nX)
data(BeckLee_mat99)
dim(BeckLee_mat99)
head(BeckLee_mat99[,1:5], 2)
tail(BeckLee_mat99[,1:5], 2)

## A list of first and last occurrences data for some fossils
data(BeckLee_ages)
head(BeckLee_ages)

## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree) ; nodelabels(cex=0.8)
@

@@

<<Setup the data>>=
## My ordinated matrix with tips only
MyData_mat50 <- BeckLee_mat50

## My ordinated matrix with tips and nodes
MyData_mat99 <- BeckLee_mat99

## My first and last occurence data
MyData_ages <- BeckLee_ages

## My phylogenetic tree
MyData_tree <- BeckLee_tree
@

@@

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(MyData_mat99, MyData_tree, method = "continuous",
    model = "acctran", time = 5, FADLAD = MyData_ages)

## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)

## Calculating disparity
sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))

## Summarising the results
summary(sum_of_ranges)
plot(sum_of_ranges, type = "continuous")
@

@@

<<dispRity class1, eval=FALSE>>=
str(sum_of_ranges)
## That's a 2512 lines output!
@

@@

<<dispRity class2>>=
## Which class is the sum_of_ranges object?
class(sum_of_ranges)

## We can summarise it using the S3 method print.dispRity
sum_of_ranges
## Some attentive ape package users will note the similarities with a phylo object...
@

@@

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem
crown <- extract.clade(MyData_tree, node = 71)$tip.label
stem <- drop.tip(MyData_tree, tip = crown)$tip.label

## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem))))

## We then can use the customised series function
crown_stem <- cust.series(MyData_mat50, factors)

## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem

## This object contains three elements
names(crown_stem)

## With "data" being the list sub-matrices
str(crown_stem$data)

## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)

## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

@@

<<discrete.series>>=
## Let's generate three time bins containing the data present every 40 Ma
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0))

## Note that the function provides a warning saying that tips where single
## points in time (no FAD/LAD information). We can fix that by adding the
## age data for the taxa that have some longer occurrence spans.
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = MyData_ages)

## To entirely avoid the warning we could collect the occurrence span data
## for all the taxa but that's not necessary. The function automatically
## assumes no occurrence span time (i.e. single points in time) for all taxa
## by default.

time_bins
@

@@

<<discrete.series2>>=
time.series(data = MyData_mat50, tree = MyData_tree, method = "discrete", time = 3)
@

@@

<<continuous.series>>=
## Let's generate four time slices every 40 Ma assuming a gradual evolution model
time_slices <- time.series(data = MyData_mat99, tree = MyData_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = MyData_ages)

time_slices

## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
    model = "gradual", time = 4)
## But that gives the same silly results (slicing every 44.50368 Ma)
@

@@

<<boot.matrix1>>=
boot.matrix(data = MyData_mat50)
## As we can wee, the output is also a dispRity object that is summarised
## automatically, and gives information on the data as well as the number of
## bootstraps and the bootstraps methods.
@

@@

<<boot.matrix2>>=
## Bootstrapping with the single bootstrap method
boot.matrix(MyData_mat50, boot.type = "single")
## Bootstrapping with the full rarefaction
boot.matrix(MyData_mat50, bootstraps = 20, rarefaction = TRUE)
## Or with a set number of rarefaction levels
boot.matrix(MyData_mat50, bootstraps = 20, rarefaction = c(6:8,3))
## And removing the last axis (default)
boot.matrix(MyData_mat50, rm.last.axis = TRUE)
## Or with a fix value (50%)
boot.matrix(MyData_mat50, rm.last.axis = 0.5)
@

@@

<<boot.matrix3>>=
## Let's do a full rarefaction on the crown/stem data for later
crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
## And just bootstrapping for the time binning/slicing
time_binsBS <- boot.matrix(time_bins)
time_slicesBS <- boot.matrix(time_slices)

## And just to remind us what's in each objects
crown_stemBS
time_binsBS
time_slicesBS
@
